Towards a miniKanren with fair search strategies

### 2019-06-25 Paper/paper.tex
ack and BFS
more ack
camera-readify
use abbrev

\documentclass[acmlarge, review=true]{acmart}
%TODO 1. delete `, review=true' ; 2. delete this line .
\usepackage[
    type={CC},       
    modifier={by-sa},
    version={4.0},   
]{doclicense}

@@

% We describe fairness levels in disjunction and conjunction
% implementations.  Specifically, a disjunction implementation can be
% fair, almost-fair, or unfair. And a conjunction implementation can
% be fair or unfair.  We compare the fairness level of four search
% strategies: the standard miniKanren interleaving depth-first search (\DFSi),
% the balanced interleaving depth-first search (\DFSbi), the fair depth-first
% search (\DFSf), and the standard breadth-first search (\BFSser).
% \DFSbi{} and \DFSf{} are new. And we present a new, more efficient
% and simpler implementation of \BFSser. Using quantitative evaluation, 
% we argue that
% \DFSbi{} and \DFSf{} are competitive alternatives to \DFSi, 
% and that our \BFSser{} implementation is more efficient than the current one.

% The syntax of a programming language should reflect its semantics. When 
%writing 
% a \conde{} expression in miniKanren, a programmer would expect all clauses 
% share the same chance of being explored, as these clauses are written in 
% parallel. The existing search strategy, interleaving depth-first search 
% (\DFSi{}), 
% however, prioritizes its clauses by the order how they are written down. 
% Similarly, when a \conde{} is followed by another goal conjunctively, a 
% programmer would expect states in parallel share the same chance of being 
% explored. Again, the answers by \DFSi{} is different from the 
% expectation. We have devised three new search strategies that have different 
% level of fairness in \disj{} and \conj{}.
%

@@

\maketitle
%\thispagestyle{empty}
%TODO uncomment the line above to be camera-ready

@@

miniKanren is a family of relational programming languages.
\citet{Friedman:2005:RS:1121583,friedman_reasoned_2018} introduce
miniKanren and its implementation in \emph{The Reasoned Schemer} 
and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). \mbox{\citet{Hemann_2016}}
describe microKanren, a minimal core of miniKanren comprised of only 54 LOC, 
miniKanren has been implemented in many other languages, including
multiple ones using the same language (e.g. OCanren\citep{kosarev2018typed}). 
% \citet{byrd2017unified} have demonstrated that miniKanren problems that at 
% first reading seem impossible are indeed possible by using a Scheme interpreter 
% written in miniKanren, thereby having access to logic variables within Scheme.
As demonstrated in \citet{byrd2017unified}, miniKanren can be used to
naturally express difficult computations, such as using an interpreter
to perform example-based program synthesis, or using a proof checker
as a theorem prover.
% miniKanren.org contains links to many miniKanren implementations and the seeds 
% of many unusual problems and their solutions in miniKanren.
The papers, talks, and tutorials on \url{miniKanren.org} present many
other unusual problems, and their solutions in miniKanren.

%miniKanren is a family of relational programming languages.
%\citet{Friedman:2005:RS:1121583,friedman_reasoned_2018} introduce
%miniKanren and its implementation in \emph{The Reasoned Schemer} 
%and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). Although originally written in 
%Scheme, miniKanren has been implemented in many other languages, including
%multiple ones using the same language. (See miniKanren.org.) This has come
%about because of \citet{Hemann_2016}'s work on microKanren, a core of 
%miniKanren comprised of only 54 LOC. Also, miniKanren
%has been ported to Ocaml, and named OCanren\citep{kosarev2018typed}). 
%\citet{byrd2017unified} have 
%demonstrated that miniKanren problems that at first reading seem impossible
%are indeed possible by using a Scheme interpreter written in miniKanren, 
%thereby
%having access to logic variables within Scheme.
%miniKanren.org contains the seeds of many interesting problems and 
%their solutions.

@@

%In the example above, the letters \texttt{c} and \texttt{d} are 
%allocated with less resources than \texttt{a}, and thus \texttt{c} and 
%\texttt{d} barely contribute to the result.

@@

% A little miniKanrener, however, may beg to differ--understanding
% implementation details and fiddling with clause order is not the first
% priority of a beginner.

@@

%To summarize our contributions, we
%\begin{itemize}
%	\item propose and implement \textbf{b}alanced \textbf{i}nterleaving 
%depth-first search (\DFSbi{}), a new search strategy with almost-fair \disj{}.
%	\item propose and implement \textbf{f}air depth-first search (\DFSf{}), 
%a new search strategy with fair \disj{}.
%	\item implement in a new way the standard breath-first search (\BFS) by 
%\citet{seres1999algebra}, a search strategy with fair \disj{} and fair 
%\conj{}. 
%We refer to \BFSser{} as their implementation and \BFSimp{} as our new 
%one. We formally prove that the two implementations are 
%semantically equivalent, however, \BFSimp{} runs faster in all benchmarks and 
%is shorter.
%\end{itemize}

Our research compares four search strategies with different features in fairness
(\autoref{table-fairness}). To summarize our contributions, we
\begin{itemize}
	\item propose and implement \textbf{b}alanced \textbf{i}nterleaving 
	depth-first search (\DFSbi{})
	\item propose and implement \textbf{f}air depth-first search (\DFSf{})
	\item implement in a new way the standard breath-first search. 
	We refer to \BFSser{} as the original implementation by 
	\citet{seres1999algebra} and \BFSimp{} as our new 
	one. We formally prove that the two implementations are 
	semantically equivalent, however, \BFSimp{} runs faster in all benchmarks 
	and 
	is shorter.
\end{itemize}

\begin{table}[h]
	\begin{tabular}{|c|c|c|}
		\hline 
		Search Strategies & disj & conj \\ 
		\hline 
		\DFSi & unfair & unfair \\ 
		\hline 
		\DFSbi & almost-fair & unfair \\ 
		\hline 
		\DFSf & fair & unfair \\ 
		\hline 
		\BFSimp, \BFSser & fair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{Fairness of all search strategies}
	\label{table-fairness}
\end{table}

@@

Now we elaborate fairness by running more queries about \repeato{}. We never 
use \runstar{} here because fairness is more interesting when we 
ask a bounded number of answers. It is perfectly fine, however, to use 
\runstar{} with any search strategy.

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to other strategies. TRS2,
chapter 10 and the appendix, ``Connecting the wires'', 
provide a comprehensive description of the miniKanren 
implementation but limited to unification constraints ($\equiv$).
Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search strategies. We follow some conventions to name variables: 
\texttt{s}s name states; \texttt{g}s (possibly with subscript) name goals; 
variables ending with $^\infty$ name spaces. Fig.~\ref{DFSi-0} shows 
the 
implementation of \disj. The 
first function, \disjtwo, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting spaces with \appendInf{}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \conjtwo{}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting space. The helper function 
\appendMapInf{} applies its input goal to states 
in its input space and composes the resulting spaces. It reuses 
\appendInf{} for space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

The strategy \textbf{b}alanced \textbf{i}nterleaving DFS (\DFSbi{}) has an 
almost-fair \disj{} and unfair 
\conj{}. The implementation of \DFSbi{} differs from 
\DFSi{}'s in the \disj{} macro. When there are one or more disjunctive 
goals, the new \disj{} builds a balanced binary tree whose leaves are the goals 
and 
whose nodes are \disjtwo{}s, hence the name of this search strategy. 
In contrast, the \disj{} in \DFSi{} constructs the binary tree in a 
particularly unbalanced form.
We list the new \disj{} with its helper in Fig.~\ref{balanced-disj}.
The new helper, \texttt{disj+}, takes two additional `arguments'. They 
accumulate goals to be put in the left and right subtrees. The first clause 
handles the case where there is only one goal. In this case, the tree is the 
goal itself. When there are more goals, we partition the list of goals 
into two sublists of roughly equal lengths and recur on the two sublists. We 
move goals to the accumulators in the last clause. As we are moving 
two goals each time, there are two base cases: (1) no goal remains; (2) one 
goal remains. We handle these two new base cases in the second clause and the 
third clause, respectively.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\disjtwo{} (Fig.~\ref{fDFS}). The new \disjtwo{} calls a new and 
fair version of \appendInf{}. \appendInfFair{} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, initialized to 
\texttt{\#{}t}, which indicates that we haven't swapped
\sInf{} and \tInf{}. The swapping 
happens at 
the third \texttt{cond} clause in \texttt{loop}, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \sInf{} and 
\tInf{} end with thunks, and that we have swapped them. In 
this case, we construct a new thunk. The new thunk swaps back the two spaces in 
the
recursive call to \texttt{loop}. This is unnecessary for fairness---we do it to 
produce answers in a more readable order.

@@

We devise two new search strategies (i.e., balanced interleaving DFS 
(\DFSbi{}) and fair DFS (\DFSf{})) and devise a new 
implementation of \BFS{} (\BFSimp{}). These strategies have different features 
in fairness: \DFSbi{} has an almost-fair \disj{} and unfair \conj{}. 
\DFSf{} has fair \disj{} and unfair \conj{}. \BFS{} has both fair
\disj{} and fair \conj{}. No search strategy here combines unfair 
\disj{} and fair \conj{}. This is because we haven't seen a case where this 
kind of search strategies would be interesting.

@@

>>>>

\documentclass[acmlarge]{acmart}
\usepackage[
    type={CC},       
    modifier={by-sa},
    version={4.0},   
]{doclicense}

@@

@@

\maketitle
\thispagestyle{empty}

@@

miniKanren is a family of relational programming languages.
\citet{Friedman:2005:RS:1121583,friedman_reasoned_2018} introduce
miniKanren and its implementation in \emph{The Reasoned Schemer} 
and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). \mbox{\citet{Hemann_2016}}
describe microKanren, a minimal core of miniKanren comprised of only 54 LOC, 
miniKanren has been implemented in many other languages, including
multiple ones using the same language (e.g. OCanren\citep{kosarev2018typed}). 
As demonstrated in \citet{byrd2017unified}, miniKanren can be used to
naturally express difficult computations, such as using an interpreter
to perform example-based program synthesis, or using a proof checker
as a theorem prover.
The papers, talks, and tutorials on \url{miniKanren.org} present many
other unusual problems, and their solutions in miniKanren.

@@

@@

@@

Our research compares four search strategies with different features in fairness
(\autoref{table-fairness}). To summarize our contributions, we
\begin{itemize}
	\item propose and implement \textbf{b}alanced \textbf{i}nterleaving 
	depth-first search (\DFSbi{})
	\item propose and implement \textbf{f}air depth-first search (\DFSf{})
	\item implement in a new way the standard breath-first search. 
	We refer to \BFSser{} as the original implementation by 
	\citet{seres1999algebra} and \BFSimp{} as our new 
	one. When we use BFS without subscripts, we mean both \BFSser{} and 
	\BFSimp{}.
	We formally prove that the two implementations are 
	semantically equivalent, however, \BFSimp{} runs faster in all benchmarks 
	and 
	is shorter.
\end{itemize}

\begin{table}[h]
	\begin{tabular}{|c|c|c|}
		\hline 
		Search Strategies & disj & conj \\ 
		\hline 
		\DFSi & unfair & unfair \\ 
		\hline 
		\DFSbi & almost-fair & unfair \\ 
		\hline 
		\DFSf & fair & unfair \\ 
		\hline 
		\BFS & fair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{Fairness of all search strategies}
	\label{table-fairness}
\end{table}

@@

Now we elaborate fairness by running more queries about \repeato{}. We never 
use \runstar{} here because fairness is more interesting when we 
ask for a bounded number of answers. It is perfectly fine, however, to use 
\runstar{} with any search strategy.

@@

In this section, we review the implementation of \DFSi{}. We focus on parts 
that are relevant to other strategies. TRS2,
chapter 10 and the appendix, ``Connecting the wires'', 
provide a comprehensive description of the miniKanren 
implementation but limited to unification constraints ($\equiv$).
Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search strategies. We follow some conventions to name variables: 
\texttt{s}s name states; \texttt{g}s (possibly with subscript) name goals; 
variables ending with $^\infty$ name spaces. Fig.~\ref{DFSi-0} shows 
the 
implementation of \disj. The 
first function, \disjtwo, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting spaces with \appendInf{}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \conjtwo{}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting space. The helper function 
\appendMapInf{} applies its input goal to states 
in its input space and composes the resulting spaces. It reuses 
\appendInf{} for space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

\DFSbi{} has almost-fair \disj{} and unfair \conj{}. Its 
implementation differs from 
\DFSi{}'s in the \disj{} macro. When there are one or more disjunctive 
goals, the new \disj{} builds a balanced binary tree whose leaves are the goals 
and 
whose nodes are \disjtwo{}s, hence the name of this search strategy. 
In contrast, the \disj{} in \DFSi{} constructs the binary tree in a 
particularly unbalanced form.
We list the new \disj{} with its helper in Fig.~\ref{balanced-disj}.
The new helper, \texttt{disj+}, takes two additional `arguments'. They 
accumulate goals to be put in the left and right subtrees. The first clause 
handles the case where there is only one goal. In this case, the tree is the 
goal itself. When there are more goals, we partition the list of goals 
into two sublists of roughly equal lengths and recur on the two sublists. We 
move goals to the accumulators in the last clause. As we are moving 
two goals each time, there are two base cases: (1) no goal remains; (2) one 
goal remains. We handle these two new base cases in the second clause and the 
third clause, respectively.

@@

\DFSf{} has fair \disj{} and unfair \conj{}. Its 
implementation differs from \DFSi{}'s in 
\disjtwo{} (Fig.~\ref{fDFS}). The new \disjtwo{} calls a new and 
fair version of \appendInf{}. \appendInfFair{} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, initialized to 
\texttt{\#{}t}, which indicates that we haven't swapped
\sInf{} and \tInf{}. The swapping 
happens at 
the third \texttt{cond} clause in \texttt{loop}, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \sInf{} and 
\tInf{} end with thunks, and that we have swapped them. In 
this case, we construct a new thunk. The new thunk swaps back the two spaces in 
the
recursive call to \texttt{loop}. This is unnecessary for fairness---we do it to 
produce answers in a more readable order.

@@

We devise two new search strategies (i.e., balanced interleaving DFS 
(\DFSbi{}) and fair DFS (\DFSf{})) and devise a new 
implementation of \BFS{} (\BFSimp{}). These strategies have different features 
in fairness: \DFSbi{} has an almost-fair \disj{} and unfair \conj{}. 
\DFSf{} has fair \disj{} and unfair \conj{}. \BFS{} has both fair
\disj{} and fair \conj{}. No search strategy here combines unfair 
\disj{} and fair \conj{}. This is because we haven't seen a case where these 
kinds of search strategies would be interesting.

@@

We thank the program committee for their insightful observations. We also thank 
our reviewers, both known and anonymous, for their corrections and suggestions.
